---
title: "Assignment 1: Introductory Statistics and Data Handling"
format: html
editor: visual
---

# Part A: Indexing, Subsetting, and Defensive R

## 1. Indexing with missing values

```{r}
# Defining the initial vector
x <- c(3.1, NA, 4.7, 2.9, NA, 5.0)
# Extracting the positions (indices) of the missing values.
na_positions <- which(is.na(x))
cat("Indices of missing values:\n")
print(na_positions)
# Extracting all non-missing values.
non_missing_x <- x[!is.na(x)]
cat("\nNon-missing values:\n")
print(non_missing_x)
# Replacing the missing values with the mean of non-missing values.
mean_non_missing <- mean(x, na.rm = TRUE)

# Assigning calculated mean to NA positions
x[is.na(x)] <- mean_non_missing
cat("\nVector after replacing NAs with the mean:\n")
print(x)
final_mean <- mean(x)

cat("\nThe new mean of x is:\n")
print(final_mean)
```

## 2. Row/column selection with multiple conditions

```{r}
# Defining the initial data frame
df <- data.frame(
  id = 1:8,
  site = c("A", "A", "B", "B", "A", "C", "C", "B"),
  treatment = c("control", "drug", "control", "drug", "drug", "control", "drug", "control"),
  score = c(12, 15, 9, 18, 14, 11, 20, 10)
)

# Subsetting the rows where site is A or C.
subset_A_or_C <- df[df$site %in% c("A", "C"), ]
cat("Rows where site is A or C:\n")
print(subset_A_or_C)
# Subsetting rows where treatment is drug and score is at least 15.
subset_drug_high_score <- df[df$treatment == "drug" & df$score >= 15, ]
cat("\nRows where treatment is drug and score >= 15:\n")
print(subset_drug_high_score)
# Creating a new object containing only columns id, treatment, and score.
df_reduced <- df[, c("id", "treatment", "score")]
cat("\nData frame with only id, treatment, and score columns:\n")
print(df_reduced)

```

Explanation: The symbol & represents logical AND and both conditions have to be TRUE, while \| represents logical OR meaning just one need to be TRUE.

## 3. Catching silent mistakes

```{r}
# Defining the initial character vector
vals <- c("1", "2", "3", "4")
# Checking whether vals is numeric before computing the mean.
is_val_numeric <- is.numeric(vals)
cat("Is 'vals' strictly numeric before conversion?", is_val_numeric, "\n")
# Condition: If it is not numeric, convert it safely.
# If FALSE, as.numeric() is used to safely cast the strings into numbers.
if (!is_val_numeric) {
  vals <- as.numeric(vals)
  cat("Converted 'vals' into a numeric vector.\n")
}
# computing the mean.
computed_mean <- mean(vals)
cat("The computed mean of the numeric vector is:", computed_mean, "\n\n")
```

Explanation: Silent coercion automatically converts data types without raising explicit errors or warnings during code execution. This hidden behaviour then leads to missing values associated with unrecognised alphabetic symbols within character strings on a routine basis. As a result, other analytical functions will compute the incorrect summary metrics, effectively spoiling the output of downstream statistical modelling completely.

# Part B: Data Types, Factors, and Common Pitfalls

## 4. Factors: ordering and reference levels

```{r}
# Defining the initial character vector
grades <- c("merit", "pass", "distinction", "pass", "merit", "pass")
# Converting the grades to a factor with ordered levels.
grades_ordered <- factor(grades, 
                         levels = c("pass", "merit", "distinction"), 
                         ordered = TRUE)
# Displaying the levels and confirming that it is ordered.
cat("Factor levels:\n")
print(levels(grades_ordered))
cat("\nIs the factor ordered?\n")
print(is.ordered(grades_ordered))
# Computing and printing a frequency table.
cat("\nFrequency table of ordered grades:\n")
print(table(grades_ordered))
```

Explanation: Ordered factors guarantee that graphical visualisation will display categorical variables in a consistent and interpretable order. Defaulting hierarchical connotations do not escape even statistical software on categorical variables, which commonly treat alphabetical text strings the same except there are little if any conventions with real ordering. Additionally, a baseline reference category is intrinsically important for interpreting linear regression coefficients.

## 5. Date parsing and type checking

```{r}
# Defining the initial character vector
dates_chr <- c("2026-01-05", "2026-01-12", "2026-02-01", "not_a_date")
# Converting the character vector into a Date vector.
dates <- suppressWarnings(as.Date(dates_chr))
cat("onverted Date vector:\n")
print(dates)
# Identifying which element(s) failed to parse.
failed_positions <- which(is.na(dates))
failed_strings <- dates_chr[failed_positions]
cat("\nElement(s) that failed to parse:\n")
print(failed_strings)
# Removing the invalid dates and computing the number of days 
# between the earliest and latest valid date.
valid_dates <- dates[!is.na(dates)]
# Subtracting the minimum date from the maximum date.
date_difference <- as.numeric(max(valid_dates) - min(valid_dates))
cat("\nNumber of days between earliest and latest valid date:\n")
print(date_difference)
```

Explanation: Standard date parsing functions operate under the assumption that character formats are strictly structured character formats, if these are to successfully perform mathematical time transformations. Standard analytical software has rigid structural requirements for conversion algorithms and breaks when encountering irregular textual inputs. This is why automated software programmes will create missing values instead of crashing a currently running data processing pipeline into pieces entirely.

## 6. Coercion with mixed types in a data frame

```{r}
# Defining the initial data frame
messy <- data.frame(
  sample_id = c("S1", "S2", "S3", "S4"),
  nitrate = c("12.1", "11.8", "missing", "13.0"),
  ph = c(6.2, 6.5, 6.1, 6.4),
  stringsAsFactors = FALSE
)
# Converting the nitrate to numeric.
messy$nitrate <- suppressWarnings(as.numeric(messy$nitrate))
cat("The 'nitrate' column has been converted to numeric:\n")
print(messy$nitrate)
# Reporting the number of NAs created by conversion.
na_count <- sum(is.na(messy$nitrate))
cat("\nNumber of NAs created by the conversion process:", na_count, "\n")
# Creating a new column nitrate_flag.
messy$nitrate_flag <- is.na(messy$nitrate)
cat("\nData frame updated with the 'nitrate_flag' column:\n")
print(messy)
# Computing the mean nitrate ignoring NAs.
mean_nitrate <- mean(messy$nitrate, na.rm = TRUE)
cat("\nThe mean nitrate value (ignoring NAs) is:", mean_nitrate, "\n")
```

# Part C: Cleaning Names, Reshaping, and Summaries

## 7. Cleaning messy column names using regex (base R only)

```{r}
# Defining the initial character vector
nm <- c(" Plot ID ", "Yield(kg/ha)", "Soil%Moisture", "2nd reading", "sensor.ID")
# Trimning the spaces.
nm_clean <- trimws(nm)
# Converting to lowercase.
nm_clean <- tolower(nm_clean)
# Replacing the punctuation and spaces with _
nm_clean <- gsub("[[:punct:] ]", "_", nm_clean)
# Collapsing the multiple _ to a single _
nm_clean <- gsub("_+", "_", nm_clean)
# Ensuring that the names do not begin with a digit with prefix with x_ if needed.
nm_clean <- sub("^([0-9])", "x_\\1", nm_clean)
# Returning the nm_clean.
cat("Cleaned column names:\n")
print(nm_clean)
```

Explanation: The first regular expression matches whitespace as well as standard punctuation marks for direct inclusion. This is followed by a greedy quantifier which detects repeated underscore symbols and condenses unnecessary sequences of characters. Lastly, the caret symbol matches the start of a number and puts a definitive restriction that the header of a variable has to start with an alphabet.

## 8. Wide → long (base R only), then grouped summaries

```{r}
# Setting the seed for reproducibility
set.seed(21)
# Defining the initial wide data frame
wide <- data.frame(
  plot = paste0("P", 1:6),
  y_2024 = rnorm(6, 5.0, 0.6),
  y_2025 = rnorm(6, 5.4, 0.6),
  y_2026 = rnorm(6, 5.2, 0.6)
)
# Reshaping the wide into long format.
long <- reshape(
  wide,
  direction = "long",
  varying = c("y_2024", "y_2025", "y_2026"),
  v.names = "yield",
  timevar = "year",
  times = c("2024", "2025", "2026"),
  idvar = "plot"
)
# Striping the default row names to provide a clean data frame
row.names(long) <- NULL
cat("Reshaped long data frame:\n")
print(head(long))
# Computing the mean and sd of yield per year.
mean_yield <- aggregate(yield ~ year, data = long, FUN = mean)
sd_yield   <- aggregate(yield ~ year, data = long, FUN = sd)
# Merging the two summary data frames together by the 'year' column.
summary_stats <- merge(mean_yield, sd_yield, by = "year")
# Renaming the columns for absolute clarity
names(summary_stats) <- c("year", "mean_yield", "sd_yield")
cat("\nSummary statistics (mean and SD) per year:\n")
print(summary_stats)
```

Explanation: The modern relational database architecture fits perfectly with long format dataset very efficiently. Statistical modelling functions often expect single outcome variables matched against specifically designed categorical predictor columns. In addition, sophisticated graphical displays require uniform mapping of aesthetic properties to specific facets of observation.

## 9. Long → wide with two value columns (tidyverse allowed)

```{r}
# Loading the package
library(tidyr)
# Defining the initial long data frame
long2 <- data.frame(
  id = rep(1:3, each = 2),
  time = rep(c("pre", "post"), times = 3),
  height = c(10, 12, 9, 11, 13, 14),
  weight = c(20, 22, 19, 21, 23, 24)
)
# Pivoting the long2 to wide format.
wide2 <- pivot_wider(
  data = long2,
  names_from = time,
  values_from = c(height, weight)
)
cat("Reshaped wide data frame:\n")
print(wide2)
# Verifying the result using str().
cat("\nStructure of the wide data frame:\n")
str(wide2)
# Computing the change scores.
wide2$height_change <- wide2$height_post - wide2$height_pre
wide2$weight_change <- wide2$weight_post - wide2$weight_pre
cat("\nData frame updated with computed change scores:\n")
print(wide2)
```

# Part D: Joins, Keys, and Debugging

## 10. Fixing a join with inconsistent keys

```{r}
# Defining the initial data frames
left_tbl <- data.frame(
  plot = c("P-01", "P-02", "P-03", "P-04"),
  yield = c(5.1, 5.4, 5.0, 5.6),
  stringsAsFactors = FALSE
)
right_tbl <- data.frame(
  plot = c("P01", "P02", "P03", "P05"),
  soil = c("clay", "silt", "sand", "peat"),
  stringsAsFactors = FALSE
)
# Standardising the plot IDs so the join works.
left_tbl$plot <- gsub("-", "", left_tbl$plot)
cat("Standardised left table keys:\n")
print(left_tbl$plot)
# Merging the tables and keep only matching rows.
merged_tbl <- merge(left_tbl, right_tbl, by = "plot")
cat("\nMerged data frame (inner join):\n")
print(merged_tbl)
# Identifying which plots were missing from each side.
missing_from_right <- setdiff(left_tbl$plot, right_tbl$plot)
missing_from_left  <- setdiff(right_tbl$plot, left_tbl$plot)
cat("\nPlots in left_tbl but missing from right_tbl:\n")
print(missing_from_right)
cat("\nPlots in right_tbl but missing from left_tbl:\n")
print(missing_from_left)
```

Explanation: Relational databases cannot match matching records across two different datasets from a merge operation because of different identifiers. This leads to standard analytical software to silently drop valuable rows of observation whenever structural links between data are disrupted by typing errors. This type of loss of information results in very biased statistical approximations in future mathematical modelling systems.

## 11. Duplicate keys and unexpected row counts

```{r}
# Defining the initial data frames
a <- data.frame(id = c(1, 2, 2, 3), score = c(10, 12, 13, 9))
b <- data.frame(id = c(1, 2, 3, 3), group = c("X", "Y", "Y", "Z"))
# Merging a and b by id.
merged_ab <- merge(a, b, by = "id")
cat("Merged data frame with duplicate keys:\n")
print(merged_ab)
# Reporting the number of rows in the merged output.

row_count <- nrow(merged_ab)
cat("\nThe merged output contains", row_count, "rows.\n\n")
```

Explanation: Cartesian products are applied automatically by relational database software when there are duplicates in the numerical value of matching identifier variables. Thus, combining algorithms yield many rows of combination data for every unique pairing between two related datasets. This type of mathematical expansion generates new records for every additional calculation, at times producing inflated total counts from original dimensions during normal data calculation routines.

## 12. Plot coloured by a derived grouping

```{r}
# Creating a version of a and b with unique IDs and merge again.
a_unique <- a[!duplicated(a$id), ]
b_unique <- b[!duplicated(b$id), ]
# Merging the deduplicated datasets.
merged_unique <- merge(a_unique, b_unique, by = "id")
cat("\n4. Merged data frame with unique IDs:\n")
print(merged_unique)
# Creating a data frame 'labels' with car names and 'eff_band'.
labels <- data.frame(
  car_name = rownames(mtcars),
  eff_band = ifelse(mtcars$mpg >= 20, "efficient", "inefficient"),
  stringsAsFactors = FALSE
)
# Merging mtcars with labels by car name (row names).
mtcars_merged <- merge(mtcars, labels, by.x = "row.names", by.y = "car_name")
# Creating a scatter plot of wt vs mpg coloured by eff_band.
point_colours <- ifelse(mtcars_merged$eff_band == "efficient", "blue", "red")
# Generating the scatter plot
plot(mtcars_merged$wt, mtcars_merged$mpg,
     col = point_colours,
     pch = 16,             # Solid filled circles for better visibility
     main = "Car Weight vs Fuel Efficiency",
     xlab = "Weight (1000 lbs)",
     ylab = "Miles Per Gallon (mpg)")
# Adding a legend with exactly two entries.
legend("topright",
       legend = c("efficient", "inefficient"),
       col = c("blue", "red"),
       pch = 16,
       title = "Efficiency Band")
```

# Part E: Functions, Lists, and Structures

## 13. A robust summary function with input checks

```{r}
# Defining the robust summary function
summarise_cols <- function(df, cols) {
  # Checking for Error with a clear message if any column name is missing
  if (!all(cols %in% names(df))) {
    stop("Error: One or more requested columns are not present in the data frame.")
  }
  # Checking for Error if any requested column is not numeric
  is_numeric_check <- sapply(df[cols], is.numeric)
  if (!all(is_numeric_check)) {
    stop("Error: One or more requested columns are not numeric.")
  }
  # Computing the required metrics using lapply
  summary_list <- lapply(cols, function(col_name) {
    # Extracting the column safely using double brackets
    x <- df[[col_name]]
    # Returning a single-row data frame for this variable
    data.frame(
      n = length(x),
      mean = mean(x, na.rm = TRUE),
      sd = sd(x, na.rm = TRUE),
      na_count = sum(is.na(x)),
      row.names = col_name
    )
  })
  # Combine the list of single-row data frames into one data frame
  result_df <- do.call(rbind, summary_list)
  return(result_df)
}
# Demonstrating the function on mtcars
cat("Summary of mtcars for mpg, hp, and wt:\n")
mtcars_summary <- summarise_cols(mtcars, c("mpg", "hp", "wt"))
print(mtcars_summary)
```

Explanation: The double bracket operator that wants to evaluate programatic character strings dynamically during automated software execution processes works correctly By contrast, the \$s symbol only works when the syntax of the literal name is explicitly typed when extracting a column in tabular form. Thus, safe usage of double brackets prevents such surprises in case analytical variables are left as such until runtime.

## 14. Applying a function over a list of data frames

```{r}
# Defining the initial list of data frames.
d1 <- iris[1:50, c("Sepal.Length", "Sepal.Width")]
d2 <- iris[51:100, c("Sepal.Length", "Sepal.Width")]
d3 <- iris[101:150, c("Sepal.Length", "Sepal.Width")]
L <- list(d1, d2, d3)
# Using lapply() to compute correlation and return as a numeric vector.
cor_list <- lapply(L, function(df) cor(df$Sepal.Length, df$Sepal.Width))
# Using unlist() coerces the resulting list strictly into a numeric vector.
cor_vector <- unlist(cor_list)
cat("Correlation within each list element (numeric vector):\n")
print(cor_vector)
# Combining the three data frames into one.
combined_df <- do.call(rbind, L)
cat("\nDimensions of the combined data frame:\n")
print(dim(combined_df))
# Computing the correlation again on the combined data.
combined_cor <- cor(combined_df$Sepal.Length, combined_df$Sepal.Width)
cat("\nCorrelation across the completely combined dataset:\n")
print(combined_cor)
```

Explanation: This programming function uses an entire list as individual arguments to perform a given analytical function. More specifically, this command pushes independent elements of a dataset into the row binding algorithm without pulling it manually. Thus, the structural agglomeration is efficient, notwithstanding having lots of different tabular objects within the programmatic workspace.

## 15. Matrix vs data frame behaviour

```{r}
# Defining the initial matrix
M <- matrix(1:12, nrow = 3, byrow = TRUE)
colnames(M) <- c("A", "B", "C", "D")
# Converting M to a data frame DF.
DF <- as.data.frame(M)
cat("Matrix converted to data frame DF:\n")
print(DF)
# Adding a non-numeric column 'band' as a factor.
DF$band <- factor(c("low", "high", "low"))
cat("\nData frame DF with new factor column 'band':\n")
print(DF)
# Showing that colSums(M) works, but colSums(DF) behaves differently.
cat("\ncolSums(M) executes successfully:\n")
print(colSums(M))
# printing the error message
cat("\ncolSums(DF) throws an error:\n")
df_error <- try(colSums(DF), silent = TRUE)
cat(attr(df_error, "condition")$message, "\n\n")
```

Explanation: Matrices are continuous mathematical vectors mapped against certain dimension attributes that need cast to very narrow data types. Similar structure, allows for complex algebraic algorithms to be computed in massive computational software packages. On the other hand, traditional tabular data structures are essentially independent lists that bind together mixed-information vectors. This flexible architectural design provides a framework to accommodate mixed observational variables within a single statistical modelling pipeline.

# Part F: Intro Statistics (Applied)

## 16. Sampling distribution by simulation

```{r}
# Setting the seed for reproducibility
set.seed(202)
# Defining the skewed population
pop <- rexp(5000, rate = 1/10)
# Taking the 200 samples of size 10, compute the mean each time.
means_10 <- replicate(200, mean(sample(pop, size = 10, replace = FALSE)))
# Repeating for sample size 40.
means_40 <- replicate(200, mean(sample(pop, size = 40, replace = FALSE)))
# Plotting the two histograms.
par(mfrow = c(1, 2))

hist(means_10, 
     main = "Sample Means (n = 10)", 
     xlab = "Calculated Mean", 
     col = "lightblue",
     border = "white")
hist(means_40, 
     main = "Sample Means (n = 40)", 
     xlab = "Calculated Mean", 
     col = "lightgreen",
     border = "white")
# Resetting plotting layout to default
par(mfrow = c(1, 1))
```

Explanation: Larger sample sizes have a noticeably tighter horizontal spread on the histogram across continuous values. Mathematically, with more observations, the result is that the sampling variance around the actual true underlying population average becomes smaller. Moreover, the visual shape is more obviously shifted from overt right skewness to symmetric normality. This typical mathematical phenomenon exemplifies the central limit theorem behavior very effectively in a fundamental statistical principle.

## 17. Confidence interval from scratch (no t.test())

```{r}
# Setting the seed for reproducibility
set.seed(77)
# Defining the initial numeric vector
x <- rnorm(18, mean = 5.5, sd = 1.2)
# Extracting the sample size, sample mean, and sample standard deviation.
n <- length(x)
x_bar <- mean(x)
s <- sd(x)
# Calculating the standard error of the mean.
se <- s / sqrt(n)
# Determining the degrees of freedom.
df <- n - 1
# Finding the critical t-value for a 95% confidence interval using a two-tailed method.
t_crit <- qt(0.975, df)
# Calculating the lower and upper bounds manually.
ci_lower <- x_bar - (t_crit * se)
ci_upper <- x_bar + (t_crit * se)
cat("Manually computed 95% Confidence Interval:\n")
cat("[", ci_lower, ",", ci_upper, "]\n\n")
# Verifying by running t.test(x) and comparing intervals.
test_result <- t.test(x)
cat("Confidence Interval from t.test():\n")
print(test_result$conf.int)
```

Explanation: Mathematically, smaller observational samples always produce more uncertainty about the precise estimations of true population variation. This necessitates theoretical probability distributions to have broader tails when plotted graphically to reflect such increased statistical uncertainty. In generalized critical thresholds, the conditions for coverage of the generalised intervals can be met as interval evaluations keeps the probability coverage at any sample size.

## 18. Two-sample comparison with checking assumptions (lightweight)

```{r}
# Setting the seed for reproducibility
set.seed(88)
# Defining the initial numeric vectors
A <- rnorm(20, mean = 50, sd = 6)
B <- rnorm(22, mean = 54, sd = 9)
# Creating side-by-side boxplots of A and B.
boxplot(A, B,
        names = c("Group A", "Group B"),
        main = "Boxplot Comparison of Group A and B",
        ylab = "Values",
        col = c("lightblue", "lightgreen"))
# Calculating the metrics for Group A
mean_A <- mean(A)
sd_A   <- sd(A)
# Calculating the metrics for Group B
mean_B <- mean(B)
sd_B   <- sd(B)
cat("Summary Statistics:\n")
cat("Group A - Mean:", round(mean_A, 2), " | SD:", round(sd_A, 2), "\n")
cat("Group B - Mean:", round(mean_B, 2), " | SD:", round(sd_B, 2), "\n\n")
# Computing a two-sample t-test.
ttest_result <- t.test(A, B)
cat("Two-Sample T-Test Result:\n")
print(ttest_result)
```

Explanation: Mathematically, the standard deviation for the second comparison dataset looks much larger. As a result, the idea that these independent observational groupings arise from populations that all have equal variances seems at least somewhat structurally implausible. Thus, using advanced statistical tests with correction degrees of freedom ensures fundamentally strong comparative conclusions.

## 19. Relevel a factor and interpret coeﬀicients

```{r}
# Using Relevel on Species so that "virginica" is the reference.
iris$Species <- relevel(iris$Species, ref = "virginica")
cat("The reference level is now:", levels(iris$Species)[1], "\n\n")
# Fitting the model where Sepal.Length ~ Species.
fit <- lm(Sepal.Length ~ Species, data = iris)
# Printing the coefficient table.
coef_table <- summary(fit)$coefficients
cat("Coefficient Table:\n")
print(coef_table)
```

Explanation: In order to mathematically analyse categorical predictor variables with linear regression models, the baseline category must be set. The intercept estimate therefore only indicates the average expected outcome for this selected reference category. The remaining coefficients for the regression are the estimated mathematical difference between each remaining category and the baseline. The reported numerical estimates associated with a reference level change cannot be changed without altering the statistical model fit.

## 20. A small end-to-end workflow

```{r}
# Setting the seed for reproducibility
set.seed(909)
# Defining the initial data frame
trial <- data.frame(
  plot = rep(paste0("P", 1:12), each = 2),
  treatment = rep(c("control", "treated"), times = 12),
  yield = rnorm(24, mean = 5.0, sd = 0.5)
)
# Applying the treatment effect
trial$yield[trial$treatment == "treated"] <- trial$yield[trial$treatment == "treated"] + 0.3
# Converting the treatment to a factor and set control as reference.
# to explicitly define "control" as the baseline category.
trial$treatment <- factor(trial$treatment)
trial$treatment <- relevel(trial$treatment, ref = "control")
cat("The reference level for treatment is:", levels(trial$treatment)[1], "\n\n")
# Computing the mean yield by treatment.
mean_yield <- aggregate(yield ~ treatment, data = trial, FUN = mean)
cat("Mean yield by treatment:\n")
print(mean_yield)
cat("\n")
# Plotting a boxplot of yield by treatment.
boxplot(yield ~ treatment, data = trial,
        main = "Agricultural Yield by Treatment Group",
        xlab = "Treatment Category",
        ylab = "Recorded Yield",
        col = c("lightblue", "lightcoral"))
# Fitting a linear model yield ~ treatment.
fit_trial <- lm(yield ~ treatment, data = trial)
cat("Linear Model Coefficients:\n")
print(summary(fit_trial)$coefficients)
```

Explanation: Both statistical methods compare mean differences between two independent observational categorical groups in a mathematical fashion. Linear modelling algorithms represent binary predictors with indicator variables, which allows them to compute direct group differences in means. Thus, this will lead to the same mathematical probability values when calculating the variance across different groups and determine overall statistical significance.
